Evaluating the Influence of Education on Current Unemployment Trends

### 2024-03-30 paper/paper.qmd
updated paper updated llm usage updated references updated scripts added parquet file put entire paper through grammarly

---
title: "Evaluating the Influence of Education on Current Unemployment Trends"
author: 
  - Adrian Ly
thanks: "Code and data are available at: [LINK](https://github.com/AdrianUofT/unemployment-employment-rate)"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

# Data

@@

# Model

@@

This strategy is designed to produce a model that is both explanatory and predictive, offering a current snapshot of the relationships in the data as well as a tool for future forecasting. Such a model can be utilized to guide decision-making processes, or pinpoint where interventions may be effective.

@@

# Results

@@

- **Impact of Higher Education**:
  - *Above Bachelor’s Degree*: The coefficient `r format(5.14, nsmall = 2)`*** indicates a significant and positive association with youth unemployment rates, implying potential overqualification or job market mismatches.
  - *Bachelor’s Degree*: Similarly, a coefficient of `r format(4.14, nsmall = 2)`** suggests that a bachelor's degree is associated with higher unemployment.
  - *University Degree*: With a coefficient of `r format(4.27, nsmall = 2)`**, university graduates also face higher unemployment rates.

- **Some Postsecondary Education**: A negative coefficient of `r format(-2.20, nsmall = 2)`+ for those with some postsecondary education indicates a decrease in unemployment compared to the reference group.

- **Adult Unemployment**: The significant positive coefficient `r format(2.19, nsmall = 2)`*** for adult unemployment implies a correlation with youth unemployment rates.

@@

- **Significance Levels**:
  - The asterisks (`***`, `**`, `*`) represent the level of statistical significance, with more stars denoting stronger evidence against the null hypothesis.

@@

Similar to the youth unemployment, @tbl-adult2019 and @tbl-adult2022 show the net percentage change in the adult unemployment for the years 2019-2023 and 2022-2023 respectively.

@@

Conversely, adults with education "Above bachelor's degree" experienced a decrease in unemployment rates, suggesting that higher education may offer better protection against job market downturns. This trend aligns with the expectation that higher qualifications translate to greater employment stability.

@@

# Discussion

@@

@@

However, the data also highlight that higher education does not guarantee immunity against unemployment. This could be due to a myriad of factors, including the mismatch between the skills acquired through higher education and the needs of the job market, or possibly the saturation of certain degrees in the labor force. A great example of this is the computer science job market. People trying to get into the computer science job market are faced with the harsh reality of not being able to find a job. In fact, "Computer Science Students Face a Shrinking Big Tech Job Market" (@citesource1) according to the NY Times with "Hiring in tech, information and media is at its lowest level since July 2020" (@citesource2). With these graduate students seeking to find a job, they are most likely to go into an "entry-level position", if they even exist regardless of the field they are pursuing. The emphasis here being that these entry-level positions are either not existent, or the people working these jobs are majorly over-qualified. This is not even accounting for the number of layoffs that happen on a regular basis from these major tech companies. A very recent major layoff is from Bell Communications where they layed off "hundreds of employees in 10-minute video call meetings" (@citesource3) and that "no union reps or members were allowed to unmute themselves during the call and ask anything" (@citesource4). Thus, while education is a powerful tool for improving employment prospects, it is not a panacea. There is a need for ongoing evaluation of educational curricula to ensure alignment with evolving industry demands.

@@

These observations suggest that the job market values not just education but the relevance and applicability of the skills obtained. In other words, people are granted "senorities", hence why they are seen as more valuable to a company. As economies evolve and new industries emerge, the adaptability of the workforce becomes crucial. Hence, continuous learning and skills development are imperative for both individuals and economies to thrive.

@@

The global economy is in a constant state of flux, influenced by technological change, globalization, and, more recently, by the impact of the COVID-19 pandemic. These shifts can create or eliminate entire industries, affecting employment opportunities differently across educational levels. For instance, the growing emphasis on technology might benefit those with relevant qualifications while displacing jobs in sectors less adaptable to these changes. Which leads into my next point.

@@

Furthermore, AI's contribution to economic efficiency and productivity has the potential to spur economic growth, indirectly fostering job creation with a shift in required skill-sets. The onus is on policymakers to navigate this transition, balancing the embrace of technological innovation with the need for robust support systems for the workforce. They face the challenge of creating policies that not only protect workers from the immediate risks of displacement but also encourage the development of industries poised to grow in the AI era. Ultimately, the integration of AI into the economy presents a complex narrative: while it may render some occupations obsolete, it simultaneously opens doors to new realms of employment. The key challenge—and opportunity—lies in harnessing AI as a tool to enhance human work, not replace it, by anticipating changes and preparing the workforce through education and training to thrive alongside intelligent machines.

@@

Outsourced jobs often transition from higher-cost labor markets to regions where labor costs are lower, fundamentally altering the employment landscape in both the originating and recipient countries. While outsourcing can lead to job losses in some sectors within developed economies, it simultaneously generates employment opportunities in developing regions. This redistribution of jobs raises questions about the long-term sustainability of such practices, considering the potential for economic disparity and the need for international labor standards.

@@

Looking ahead, the tension between the advantages of outsourcing and its impact on the job market will persist as a key issue. It is imperative for policy to address the repercussions for workers while supporting economic growth and maintaining competitive advantages in a globalized economy. This includes investing in education and vocational training to equip the domestic workforce with skills that are in demand and less likely to be outsourced. Additionally, there is a growing call for multinational corporations to consider the ethical implications of their outsourcing strategies and strive for practices that are equitable and sustainable across all regions.

@@

Weaknesses and next steps should also be included.

@@

>>>>

---
title: "Evaluating the Influence of Education on Current Unemployment Trends"
author: 
  - Adrian Ly
thanks: "Code and data are available at: [LINK](https://github.com/AdrianUofT/unemployment-employment-rate)"
date: today
date-format: long
abstract: "This paper investigates the relationship between educational attainment and unemployment rates amidst the evolving dynamics of technological advancement and globalization, utilizing data from the Labour Force Survey spanning 2019 to 2023. Our analysis reveals significant variability in unemployment rates across different educational levels, highlighting the nuanced impact of higher education in an era increasingly dominated by artificial intelligence and outsourcing. These findings underscore the critical need for educational systems to adapt, emphasizing skills that complement technological advancements and align with the global job market. Ultimately, this study illuminates the complex interplay between education, employment, and technological change, offering valuable insights for policymakers and educators in preparing the workforce for the challenges of the future labor market."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

In recent years, the narrative surrounding employment and the value of higher education has become increasingly complex. Amidst a landscape marked by technological advancements and a globalizing economy, stories of job seekers with advanced degrees facing employment challenges have proliferated. There are even "trends" going around social media on platforms such as TikTok or Instagram, where people are sharing their experience of graduating with a degree, in some cases multiple degrees, and still struggling to find a job or a job that they perceive their degree to be worth (i.e. working as a barista at Starbucks when they have a computer science degree, etc.). News reports often depict long queues for scarce job openings, painting a vivid picture of the competitive and sometimes disheartening job market. This can be seen from news sources such as the `Toronto Sun` (@citesource9), the `blogTO` (@citesource8), and many more. This juxtaposition of high educational attainment and employment difficulty raises critical questions about the evolving dynamics of the labor market, the role of education, and the implications for future workforce development.

This paper delves into the nuanced relationship between educational levels and unemployment rates, set against the backdrop of rapid technological change and globalization. By analyzing unemployment trends across various demographics from 2019 to 2023, this study seeks to uncover how different levels of education impact an individual's likelihood of employment. Amid widespread discourse on the diminishing returns of higher education in certain sectors, and the transformative potential of artificial intelligence (AI) and automation to replace traditional jobs, this research aims to provide empirical insights into these phenomena.

A significant gap exists in understanding the real-time implications of these global shifts on employment, particularly in how educational attainment intersects with evolving job market demands. This study addresses this gap by utilizing data from the Labor Force Survey, exploring the interplay between education and unemployment, and considering the external factors influencing these relationships.

The findings reveal varied unemployment rates across educational strata, with notable differences in how higher education correlates with job security in the face of AI integration and outsourcing trends. These results underscore the necessity for a reevaluation of educational objectives and curricula, emphasizing skills that are less susceptible to automation and more aligned with emerging industry needs.

The importance of this research lies in its potential to inform policy decisions and educational strategies, aiming to foster a workforce that is adaptable, skilled, and prepared for the challenges of the future job market. By providing a clearer understanding of the current employment landscape, this paper contributes to the broader discourse on preparing for a labor market in flux.

The paper is structured as follows: Following this introduction, we describe the methodology and data sources used in our analysis in (@sec-data). Subsequent sections present the model we estimate (@sec-model), and results (@sec-results) showcasing the implications of educational attainment on unemployment rates. The paper concludes with a discussion (@sec-discussion) of the findings, limitations, and recommendations for future research and policy considerations.

# Data {#sec-data}

@@

# Model {#sec-model}

@@

This strategy is designed to produce a model that is both explanatory and predictive, offering a current snapshot of the relationships in the data as well as a tool for future forecasting. Such a model can be utilized to guide decision-making processes or pinpoint where interventions may be effective.

@@

# Results {#sec-results}

@@

- **Impact of Higher Education**:
  - *Above Bachelor’s Degree*: The coefficient `r format(5.14, nsmall = 2)` indicates a significant and positive association with youth unemployment rates, implying potential overqualification or job market mismatches.
  - *Bachelor’s Degree*: Similarly, a coefficient of `r format(4.14, nsmall = 2)`** suggests that a bachelor's degree is associated with higher unemployment.
  - *University Degree*: With a coefficient of `r format(4.27, nsmall = 2)`, university graduates also face higher unemployment rates.

- **Some Postsecondary Education**: A negative coefficient of `r format(-2.20, nsmall = 2)` for those with some postsecondary education indicates a decrease in unemployment compared to the reference group.

- **Adult Unemployment**: The significant positive coefficient `r format(2.19, nsmall = 2)` for adult unemployment implies a correlation with youth unemployment rates.

@@

@@

Similar to youth unemployment, @tbl-adult2019, and @tbl-adult2022 show the net percentage change in adult unemployment for the years 2019-2023 and 2022-2023 respectively.

@@

Conversely, adults with education "Above a bachelor's degree" experienced a decrease in unemployment rates, suggesting that higher education may offer better protection against job market downturns. This trend aligns with the expectation that higher qualifications translate to greater employment stability.

@@

# Discussion {#sec-discussion}

@@

From the results of our analysis, two key insights emerge that expand our understanding of the global labor market and its interplay with education.

@@

However, the data also highlight that higher education does not guarantee immunity against unemployment. This could be due to a myriad of factors, including the mismatch between the skills acquired through higher education and the needs of the job market, or possibly the saturation of certain degrees in the labor force. A great example of this is the computer science job market. People trying to get into the computer science job market are faced with the harsh reality of not being able to find a job. In fact, "Computer Science Students Face a Shrinking Big Tech Job Market" (@citesource1) according to the NY Times "Hiring in tech, information, and media is at its lowest level since July 2020" (@citesource2). With these graduate students seeking to find a job, they are most likely to go into an "entry-level position", if they even exist regardless of the field they are pursuing. The emphasis here is that these entry-level positions are either not existent, or the people working these jobs are majorly over-qualified. This does not even account for the number of layoffs that happen on a regular basis from these major tech companies. A very recent major layoff is from Bell Communications where they played off "hundreds of employees in 10-minute video call meetings" (@citesource3) and that "no union reps or members were allowed to unmute themselves during the call and ask anything" (@citesource4). Thus, while education is a powerful tool for improving employment prospects, it is not a panacea. There is a need for ongoing evaluation of educational curricula to ensure alignment with evolving industry demands.

@@

These observations suggest that the job market values not just education but the relevance and applicability of the skills obtained. In other words, people are granted "seniorities", hence why they are seen as more valuable to a company. As economies evolve and new industries emerge, the adaptability of the workforce becomes crucial. Hence, continuous learning and skills development are imperative for both individuals and economies to thrive.

@@

The global economy is in a constant state of flux, influenced by technological change, globalization, and, more recently, by the impact of the COVID-19 pandemic. A recent news article publishing the tech sector layoffs from companies such as Apple, Dell, IBM, and many more states that "the tech industry’s wave of layoffs shows no signs of slowing down as we enter the latter half of March 2024. Major players across the sector are still downsizing their workforces as economic headwinds persist" (@citesource5). These economic shifts can create or eliminate entire industries, affecting employment opportunities differently across educational levels. The growing emphasis on technology might benefit those with relevant qualifications while displacing jobs in sectors less adaptable to these changes. This leads into my next point.

@@

Furthermore, AI's contribution to economic efficiency and productivity has the potential to spur economic growth, indirectly fostering job creation with a shift in required skill sets. The onus is on policymakers to navigate this transition, balancing the embrace of technological innovation with the need for robust support systems for the workforce. They face the challenge of creating policies that not only protect workers from the immediate risks of displacement but also encourage the development of industries poised to grow in the AI era. Ultimately, the integration of AI into the economy presents a complex narrative: while it may render some occupations obsolete, it simultaneously opens doors to new realms of employment. The key challenge—and opportunity—lies in harnessing AI as a tool to enhance human work, not replace it, by anticipating changes and preparing the workforce through education and training to thrive alongside intelligent machines.

@@

Outsourced jobs often transition from higher-cost labor markets to regions where labor costs are lower, fundamentally altering the employment landscape in both the originating and recipient countries. While outsourcing can lead to job losses in some sectors within developed economies, it simultaneously generates employment opportunities in developing regions. This redistribution of jobs raises questions about the long-term sustainability of such practices, considering the potential for economic disparity and the need for international labor standards. Recently, Canada Life is said to be cutting out their IT department and is instead "Outsourcing work to a tech company headquartered in India" (@citesource6), again where labor costs are much cheaper.

@@

As we interpret the results and consider their implications for the future, a central theme emerges the evolving landscape of employment in the face of advancing artificial intelligence (AI) and the globalization of the workforce. These dynamics present a dual challenge: on the one hand, the integration of AI into various sectors threatens to supplant jobs traditionally performed by humans; on the other, the trend of outsourcing work internationally poses its own set of challenges for domestic employment.

Companies are increasingly inclined to adopt AI technologies not only for their potential to cut costs and increase efficiency but also for their ability to perform tasks with precision and consistency that outmatches human capability in certain areas. This push towards automation is evident across a range of industries, from manufacturing and logistics to services and administration. The allure of AI lies in its promise of perpetual productivity without the inherent limitations of human labor, such as the need for rest, susceptibility to error, or demands for benefits and fair wages. There is now this push to "Learn AI now or risk losing your job, experts warn" (@citesource7).

Simultaneously, the global market offers opportunities for companies to outsource work to regions where labor costs are significantly lower, further exacerbating the displacement of jobs in higher-cost countries. This trend is facilitated by the digital age, which allows for seamless collaboration and communication across continents, making geographical boundaries less relevant to where and how work is performed.

These shifts necessitate a reevaluation of the role of human workers in an increasingly automated and globalized job market. The key for individuals is to cultivate skills and attributes that are less susceptible to replacement by AI or outsourcing. This includes:

- **Complex Problem-Solving**: Developing the ability to tackle intricate issues that require nuanced understanding and creative thinking, areas where AI still lags behind human capability.

- **Emotional Intelligence**: Enhancing skills related to emotional awareness, empathy, and interpersonal communication, which are crucial in professions such as healthcare, education, and customer service where human interaction is paramount.

- **Adaptability and Lifelong Learning**: Staying agile in the face of technological advancements by continuously updating one's skill set and embracing new knowledge areas, particularly those that are complementary to AI technologies.

- **Innovation and Creativity**: Leveraging innate human creativity to drive innovation, an area where AI can assist but not originate.

- **Physical Presence**: Jobs that require a physical presence as mentioned before are less likely to be replaced by AI. For instance, it is difficult for AI to replace people such as plumbers firefighters or people working in trades.

As we move forward, the interaction between human labor, AI, and the global workforce will shape the future of employment. Companies and individuals alike must navigate these waters with foresight, balancing the benefits of technological and global workforce integration with the need to maintain a vibrant, engaged, and skilled human workforce. The challenge lies not in resisting change but in adapting to it, ensuring that the workforce of the future is equipped to complement emerging technologies rather than be displaced by them.

@@

In evaluating the data and methodology utilized in our analysis, several weaknesses warrant attention, alongside considerations for future data collection efforts aimed at refining our understanding of the labor market dynamics.

One notable limitation is the potential for sampling bias inherent in survey-based data collection methods like the Labour Force Survey. Despite rigorous sampling techniques, certain segments of the population, such as transient workers or those in informal employment sectors, might be underrepresented, leading to gaps in the captured data. This underrepresentation could skew the unemployment rates and not fully reflect the nuances of the job market.

Furthermore, the data's reliance on self-reported information poses a challenge, as it may be subject to response bias. Individuals' interpretations of their employment status or educational level can vary, potentially leading to inconsistencies in the categorization and measurement of unemployment rates.

Additionally, the rapid pace of change in the modern economy, driven by factors such as technological advancements and globalization, means that data can quickly become outdated. The interval between data collection and publication, though relatively short, might not capture the most current labor market trends, especially in times of economic volatility.

To address these limitations and enhance the robustness of future data collection efforts, several steps can be taken:

- **Expanding Coverage**: Efforts should be made to include hard-to-reach populations and those in non-traditional employment arrangements to ensure a more comprehensive representation of the labor force.

- **Enhancing Data Collection Methods**: Incorporating a mix of quantitative and qualitative data collection methods could provide a more nuanced view of the labor market. This might involve detailed interviews, case studies, and the use of alternative data sources like social media analytics or real-time labor market data from online job platforms.

- **Leveraging Technology**: Adopting advanced data collection technologies, such as mobile surveys or web scraping techniques, can offer more timely and frequent labor market insights, enabling a near-real-time analysis of trends.

- **Cross-validation with Other Data Sources**: To ensure the reliability and accuracy of the data, cross-validation with other labor market studies and administrative records can help identify and rectify discrepancies or biases in the survey data.

- **Fostering International Collaboration**: As outsourcing and the global distribution of labor become increasingly prevalent, international collaboration in labor market research can provide a more holistic view of global employment trends, helping to contextualize domestic data within the broader global economy.

By addressing these weaknesses and adopting forward-looking data collection strategies, future research can provide deeper insights into the evolving labor market, ultimately informing more effective policy interventions and workforce development initiatives.

@@

In conclusion, our exploration of unemployment rates across various educational levels and demographics, set against the backdrop of an evolving technological landscape and globalized economy, reveals intricate patterns and raises critical questions about the future of work. The data suggests that educational attainment plays a significant role in employment prospects, yet it is clear that the nature of education and the fields of study are pivotal in determining its true value in the job market. As artificial intelligence and automation become increasingly prevalent, the threat of job displacement looms large, not just for manual or routine jobs but also for roles traditionally considered secure within the realms of higher education.

The trend of outsourcing further complicates the employment narrative, presenting both challenges and opportunities as companies seek to leverage the global talent pool. This shift necessitates a reevaluation of domestic workforce strategies and underscores the importance of fostering skills that are resilient to these global trends, such as emotional intelligence, creativity, and adaptability.

As we stand at the cusp of significant shifts in the labor market, policymakers, educators, and individuals must engage in proactive dialogue and action. There is a pressing need to align educational programs with future job market demands, emphasizing not just technical skills but also the soft skills that distinguish human workers from AI. Moreover, lifelong learning must be integrated into career trajectories to ensure that the workforce remains adaptable to emerging technologies and industry shifts.

The journey ahead is complex and fraught with uncertainties, yet it also offers a canvas for innovation and adaptation. By embracing change and viewing it as an opportunity for growth, we can navigate the challenges of AI integration and globalization, ensuring a future where human labor complements technological advancements, leading to a more resilient and dynamic job market.

In essence, our findings call for a collective reimagining of the future of work, where education, policy, and workforce development converge to create a labor market that is inclusive, adaptive, and forward-looking. As we move forward, it is clear that the intersection of technology, education, and labor will continue to be a critical area of study, shaping the contours of our economic and social landscape in the years to come.
